A Topic Model of Climate Change Literature

### 2019-06-26 main.tex
jan's edits

% Title
\title{A Topic Model of Climate Change Literature}
\title{Words, words, words: Mapping the Matter of Climate Change Literature}
\title{A Topography of Climate Change Research}
\author[1,2]{Max Callaghan}

\affil[1]{Mercator Research Institute on Global Commons and Climate Change, Torgauer Straße, 10829 Berlin, Germany}
\affil[2]{School of Earth and Environment, University of Leeds, Leeds LS2 9JT, United Kingdom}

@@

\noindent We live in an age of ``Big Literature'' 
\cite{Nunez-Mir2016}, where the science of climate change is expanding rapidly \cite{Grieneisen2011, Haunschild2016}. In the five years since the publication of the last IPCC assessment report \cite{IPCC2014c}, 201,606 papers were published (see Table \ref{tab}). This is almost as much as the 204,585 papers published during the first five assessment periods, a period of nearly 30 years. The vocabulary of the literature has also expanded - from 2,000 unique words in AR1 to 94,746 words so far in AR6 - indicating that the field has incorporated new content. For example, the zika virus, which was mentioned in 182 articles from 2014-2018, had never before been discussed in the titles or abstracts of articles relating to climate change. Yet it has emerged as a topic of high relevance: the incidence of the virus, the outbreak of which in Brazil in 2016 was declared a public health emergency by the WHO, is set to increase under rising global temperatures \cite{Rao2019}. Similar rapid emergences can be seen for [Intended] Nationally Determined Contributions ([I]NDCs) and Sustainable Development Goals (SDGs) in AR6, and Biochar and REDD (The United Nations Collaborative Programme on Reducing Emissions from Deforestation and Forest Degradation in Developing Countries) in AR5, among others. This large and rapidly evolving literature demands innovative machine-reading methods to enable understanding of the field at scale.

@@

Research synthesis offers structured methodologies for integrating knowledge from bodies of evidence \cite{Chalmers2002}. Systematic reviews, which are common in the health sciences, collect all literature relevant to a specific question, in order to establish whether findings are consistent and generalisable \cite{Mulrow597}. Already in 1994, systematic review was being proposed as an answer to the ``small mountain of information'' presented by two million annual publications in biomedical literature \cite{Mulrow597}. Although climate change faces a similar scale problem, systematic review is little used in the field \cite{Berrang-Ford2015}. Systematic maps are a more recent methodological development, which show the state of the evidence in a broad subject area \cite{James2016}. There have been high profile calls to broaden the applications of evidence mapping in sustainable development \cite{McKinnon2015}, and they are similar underused in climate change, with only 25 of the more than 400,000 relevant papers containing the phrase "systematic map" in the abstract or title.

In the research synthesis field, researchers are introducing new methods from natural language processing and machine-learning to scale up synthesis work \cite{OMara-Eves2015, Westgate2018a}. In the same way, this study uses topic modelling \cite{Blei2010} to map out the vast body of evidence on climate change. Topic modelling is an unsupervised machine-learning technique, where patterns of word co-occurrences in documents are used to learn a set of topics which can be used to describe the corpus. The word topic derives from the Greek word for place (topos), and by \textit{situating} the documents in a reduced-form projection of their thematic content (see Figure \ref{oecd_topic_map}), we create a \textit{topographic map} of the literature on climate change. Such a systematic engagement with the thematic content of the climate science is missing from the literature so far.%To our knowledge, this is the first

Systematic maps can have several uses: from the identification of knowledge gaps and gluts to the can identification of authors or institutions with expertise on a subject. The map in this paper also has several potential uses, but we focus here on using the map to understand the evolution of relationship between the five IPCC assessment reports and the wider literature on climate change. Big Literature in climate change poses a great challenge to the IPCC, who's mandate is to comprehensively assess the science of climate change \cite{IPCC2013}. The ratio of studies cited in IPCC reports to the number or relevant studies has declined from 60\% to 20\%  \cite{Minx2017l}, thus increasing the risk of a lack of comprehensiveness, or even bias.

What gets included in IPCC reports is also much discussed. Policymakers demand more solution-oriented knowledge \cite{Kowarsch2017}, and there have been calls for the inclusion of more knowledge from the social sciences \cite{Victor2015}. But what we say about the literature cited in the IPCC has often been limited by a failure to consider the context of the wider literature. 
A scientometric study from 2011 claimed that the IPCC gave a greater \textit{emphasis} to natural sciences and, within the social sciences, to economics \cite{Bjurström2011}. This claim has subsequently been interpreted as a disciplinary bias by the IPCC \cite{Hulme2010, Corbera2016} towards the natural sciences, but the study operationalised disciplinary emphasis as simply the share of citations from each field, and was based on analysis of the Third Assessment report, published in 2001. In the following sections, we compare distributions of literature in the subset that is cited by the IPCC and the wider corpora, to analyse the \textit{representation}, proportional or otherwise, of scientific fields, as well as of topics.

@@

Computational methods then, provide \textit{assistance} to humans in understanding a climate change literature that is produced by human individuals and institutions.  As in other areas of machine learning \cite{Barocas2016}, computational assessments of the literature run the risk of reproducing biases or gaps in the training data. However, just as with other applications of machine learning, machine-reading the science of climate change can provide information that helps us to understand gaps in what we have knowledge about, and potential biases, or options for emphasis, in how we represent that knowledge in the IPCC. In this way, we can more effectively meet the challenges posed by Big Literature to global environmental assessments.
%Machine learning approaches to understanding scientific literature are already latently present in our engagement with  when we use academic search engines or are recommended articles

@@

%\listoffigures
\linespread{1}
\bibliography{Mendeley}
\bibliographystyle{unsrt}

>>>>

% Title
\title{A Topic Model of Climate Change Literature}
\title{Words, words, words: Mapping the Matter of Climate Change Literature}
\title{A Topography of Climate Change Research}
\author[1,2]{Max Callaghan}
\author[1,2]{Jan Minx}
\author[2]{Piers M. Forster}

\affil[1]{Mercator Research Institute on Global Commons and Climate Change, Torgauer Straße, 10829 Berlin, Germany}
\affil[2]{Priestley International Centre for Climate, University of Leeds, Leeds LS2 9JT, United Kingdom}

@@

		\noindent We live in an age of ``Big Literature'' 
		\cite{Nunez-Mir2016, Minx2017}, where the science of climate change is expanding exponentially \cite{Grieneisen2011, Haunschild2016}. For example, in the five years since the publication of the last IPCC assessment report \cite{IPCC2014c}, 201,606 papers were published in WoS (see Table \ref{tab}). This is almost as much as the 204,585 papers published during the first five assessment periods, a period of nearly 30 years. A total of about 350,000 new publications can be expected for the current sixth assessment cycle of the IPCC. If we assume that WOS contains roughly about 41\% of the entire climate change literature (Khabasa), this yields a total number of new publications of around 850,000 for this assessment cycle. But the literature has also become more diverse. This is, for example, reflected in the expansion of vocabulary in this literature - from 2,000 unique words in AR1 to 94,746 words so far in AR6 - indicating that the field has incorporated new content. For example, the zika virus, which was mentioned in 182 articles from 2014-2018, had never before been discussed in the titles or abstracts of articles relating to climate change. Yet it has emerged as a topic of high relevance: the incidence of the virus, the outbreak of which in Brazil in 2016 was declared a public health emergency by the WHO, is set to increase under rising global temperatures \cite{Rao2019}. Similar rapid emergences can be seen for [Intended] Nationally Determined Contributions ([I]NDCs) and Sustainable Development Goals (SDGs) in AR6, and Biochar and REDD (The United Nations Collaborative Programme on Reducing Emissions from Deforestation and Forest Degradation in Developing Countries) in AR5, among others.

@@

		Big literature poses at least three challenges for scientific policy advice and its underlying science: \emph{First}, established procedures in scientific assessments like those conducted by the IPCC fail to address the exploding literature base. For example, the ratio of studies cited in IPCC reports to the number or relevant studies has declined from 60\% to 20\%  \cite{Minx2017l} and poses a rapidly growing risk of selection bias. More generally, the provision of comprehensive, objective, open and transparent assessments of the available scientific literature as often mandated is no longer possible for authors or author teams by traditional means. Machine reading and learning methods as well as other data science applications are required to enable an understanding of the field of climate change research at scale (REFs). \emph{Second}, evidence synthesis - the enterprise of reviewing the literature based on a formal and systematic set of methods (REFs) - becomes increasingly important for aggregating and consolidating the rapidly emerging knowledge and enabling scientific assessments to do their job. Yet, traditional methods of evidence synthesis themselves are pushed to their limits by the large amount of scientific publications, whilst the field of evidence synthesis technology that tries to streamline human tasks through  at the different stages of the review process is still in its infancy (REFs). \emph{Finally}, overwhelming amounts of literature may be a major reasons why assessments of scientific assessments (REFs) do not offer robust quantifications, for example, between report citations and the underlying literature. This may substantially bias some of the findings.

		This study uses topic modelling \cite{Blei2010} to map out the vast body of evidence on climate change. Topic modelling is an unsupervised machine-learning technique, where patterns of word co-occurrences in documents are used to learn a set of topics which can be used to describe the corpus. The word topic derives from the Greek word for place (topos), and by \textit{situating} the documents in a reduced-form projection of their thematic content (see Figure \ref{oecd_topic_map}), we create a \textit{topographic map} of the literature on climate change. Such a systematic engagement with the thematic content of the climate science is missing from the literature so far.%To our knowledge, this is the first
		We apply this map in a second step to understand how the IPCC reports have represented the available climate change literature and re-evaluate claims of bias based on a comprehensive understanding of the available climate science.

		\subsection{Mapping out the landscape of climate change literature}

		\begin{itemize}
			\item There is a broad and diverse range of topics
			\item The climate change literature is dominated by the Natural Sciences, but there is a range of interdisciplinary topics. Few topics are dominated by other discisplines.
			\item Topic growth within the landscape is dynamic. While most topics exist from the beginning, they show distinct patterhs of growth.
		\end{itemize}

		\subsection{Research representation in IPCC reports}

		\subsection{Discussion}

		\begin{itemize}
			\item Focus of our search - not a full representation - this may bias some of our results
			\item There is a much wider range of applications for this type of research
		\end{itemize}

@@

		Computational methods then, provide \textit{assistance} to humans in understanding a climate change literature that is produced by human individuals and institutions.  As in other areas of machine learning \cite{Barocas2016}, computational assessments of the literature run the risk of reproducing biases or gaps in the training data. However, just as with other applications of machine learning, machine-reading the science of climate change can provide information that helps us to understand gaps in what we have knowledge about, and potential biases, or options for emphasis, in how we represent that knowledge in the IPCC. In this way, we can more effectively meet the challenges posed by Big Literature to global environmental assessments.

@@

	%\listoffigures
	\linespread{1}
	%\bibliography{Mendeley}
	\begin{thebibliography}{10}

		\bibitem{Bjurström2011}
		Andreas Bjurstr{\"{o}}m and Merritt Polk.
		\newblock {Physical and economic bias in climate change research: A
			scientometric study of IPCC Third Assessment Report}.
		\newblock {\em Climatic Change}, 108(1):1--22, 2011.

		\bibitem{Nunez-Mir2016}
		Gabriela~C Nunez-Mir, Basil~V Iannone, Bryan~C Pijanowski, Ningning Kong,
		Songlin Fei, and Richard Fitzjohn.
		\newblock {Automated content analysis: addressing the big literature challenge
			in ecology and evolution}.
		\newblock {\em Methods in Ecology and Evolution}, 7(11):1262--1272, 2016.

		\bibitem{Grieneisen2011}
		Michael Grieneisen and Minghua Zhang.
		\newblock {The Current Status of Climate Change Research}.
		\newblock {\em Nature Climate Change}, 1:72--73, 2011.

		\bibitem{Haunschild2016}
		Robin Haunschild, Lutz Bornmann, and Werner Marx.
		\newblock {Climate Change Research in View of Bibliometrics}.
		\newblock {\em PLoS ONE}, 11(7):1--19, 2016.

		\bibitem{IPCC2014c}
		IPCC.
		\newblock {\em {Climate Change 2014: Synthesis Report. Contribution of Working
				Groups I, II and III to the Fifth Assessment Report of the Intergovernmental
				Panel on Climate Change}}.
		\newblock IPCC, Geneva, Switzerland, 2014.

		\bibitem{Rao2019}
		V.~Brahmananda Rao, K.~Maneesha, Panangipalli Sravya, Sergio~H. Franchito,
		Hariprasad Dasari, and Manoel~A. Gan.
		\newblock {Future increase in extreme El Nino events under greenhouse warming
			increases Zika virus incidence in South America}.
		\newblock {\em npj Climate and Atmospheric Science}, 2(1):2--8, 2019.

		\bibitem{Chalmers2002}
		Iain Chalmers, Larry~V Hedges, and Harris Cooper.
		\newblock {A Brief History of Research Synthesis}.
		\newblock {\em Evaluation {\&} The Health Professions}, 25(1):12--37, 2002.

		\bibitem{Mulrow597}
		C~D Mulrow.
		\newblock {Systematic Reviews: Rationale for systematic reviews}.
		\newblock {\em BMJ}, 309(6954):597--599, 1994.

		\bibitem{Berrang-Ford2015}
		Lea Berrang-Ford, Tristan Pearce, and James~D. Ford.
		\newblock {Systematic review approaches for climate change adaptation
			research}.
		\newblock {\em Regional Environmental Change}, 15(5):755--769, 2015.

		\bibitem{James2016}
		Katy~L. James, Nicola~P. Randall, and Neal~R. Haddaway.
		\newblock {A methodology for systematic mapping in environmental sciences}.
		\newblock {\em Environmental Evidence}, 5(1):1--13, 2016.

		\bibitem{McKinnon2015}
		Madeleine~C. McKinnon, Samantha~H. Cheng, Ruth Garside, Yuta~J. Masuda, and
		Daniel~C. Miller.
		\newblock {Sustainability: Map the evidence}.
		\newblock {\em Nature}, 528(7581):185--187, 2015.

		\bibitem{OMara-Eves2015}
		Alison O'Mara-Eves, James Thomas, John McNaught, Makoto Miwa, and Sophia
		Ananiadou.
		\newblock {Using text mining for study identification in systematic reviews: A
			systematic review of current approaches}.
		\newblock {\em Systematic Reviews}, 4(1):1--22, 2015.

		\bibitem{Westgate2018a}
		Martin~J. Westgate, Neal~R. Haddaway, Samantha~H. Cheng, Emma~J. McIntosh,
		Chris Marshall, and David~B. Lindenmayer.
		\newblock {Software support for environmental evidence synthesis}.
		\newblock {\em Nature Ecology and Evolution}, 2(4):588--590, 2018.

		\bibitem{Blei2010}
		David Blei, Lawrence Carin, and David Dunson.
		\newblock {Probabilistic topic models}.
		\newblock {\em IEEE Signal Processing Magazine}, 2010.

		\bibitem{IPCC2013}
		IPCC.
		\newblock {Principles governing IPCC work}, 2013.

		\bibitem{Minx2017l}
		Jan~C. Minx, Max Callaghan, William~F. Lamb, Jennifer Garard, and Ottmar
		Edenhofer.
		\newblock {Learning about climate change solutions in the IPCC and beyond}.
		\newblock {\em Environmental Science {\&} Policy}, 2017.

		\bibitem{Kowarsch2017}
		Martin Kowarsch, Jason Jabbour, Christian Flachsland, Marcel T.~J. Kok, Robert
		Watson, Peter~M. Haas, Jan~C. Minx, Joseph Alcamo, Jennifer Garard, Pauline
		Riousset, L{\'{a}}szl{\'{o}} Pint{\'{e}}r, Cameron Langford, Yulia Yamineva,
		Christoph von Stechow, Jessica O'Reilly, and Ottmar Edenhofer.
		\newblock {A road map for global environmental assessments}.
		\newblock {\em Nature Climate Change}, 7(6):379--382, 2017.

		\bibitem{Victor2015}
		{David G. Victor}.
		\newblock {Embed the social sciences in climate policy - David Victor}.
		\newblock {\em Nature}, 520:7--9, 2015.

		\bibitem{Hulme2010}
		Mike Hulme and Martin Mahony.
		\newblock {Climate change: What do we know about the IPCC?}
		\newblock {\em Progress in Physical Geography}, 34(5):705--718, 2010.

		\bibitem{Corbera2016}
		Esteve Corbera, Laura Calvet-Mir, Hannah Hughes, and Matthew Paterson.
		\newblock {Patterns of authorship in the IPCC Working Group III report}.
		\newblock {\em Nature Climate Change}, 6(1):94--99, 2016.

		\bibitem{Ford2016b}
		James~D. Ford, Laura Cameron, Jennifer Rubis, Michelle Maillet, Douglas
		Nakashima, Ashlee~Cunsolo Willox, and Tristan Pearce.
		\newblock {Including indigenous knowledge and experience in IPCC assessment
			reports}.
		\newblock {\em Nature Climate Change}, 6(4):349--353, 2016.

		\bibitem{Barocas2016}
		Solon Barocas and Andrew~D Selbst.
		\newblock {Big Data's Disparate Impact}.
		\newblock {\em California Law Review}, 104:671--732, 2016.

	\end{thebibliography}
	\bibliographystyle{unsrt}
